A* with Fast and Tight Heuristic/Potentials in Road Networks

### 2020-08-05 paper/ch_potentials.tex
foo

\pdfinfo{
/Title (A* with Perfect Heuristic/Potentials in Road Networks)
/Author (Ben Strasser, Tim Zeitz)
/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
}
\title{A* with Perfect Heuristic/Potentials in Road Networks}
\author{Ben Strasser\\
academia@ben-strasser.net\\
\And
Tim Zeitz\\
tim.zeitz@kit.edu\\
Karlsruhe Institute of Technology
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and optimal routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic, also called potential, function.
Computing a good heuristic/potential is a challenge of its own.
To avoid the need for a heuristic/potential, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
These techniques are not based on A* and therefore do not need a heuristic/potential.
They achieve speed and optimality but sacrifize A*'s flexibility.
In this paper, we use hierarchical techniques to compute heuristic/potential functions.
We call our technique CH-Potential.
Additionally, we describe some A* optimizations to accelerate the processing of low degree nodes, which occur in road networks.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
\end{abstract}

@@

In this paper, we consider an algorithm, where the query phase consists of an A* search.
The preprocessing phase computes auxilary data to efficiently evaluate the heuristic/potential.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
Following, their terminology, we refer to the heuristic as potential function $\pi$.
Further, paths are computed from a source node $s$ to a target node $t$.

We say that a potential $\pi$ is \emph{perfect}, if no tighter potential can be computed in the preprocessing phase.
The shortest path distances with respect to $w_\ell$ are perfect.
CH-Potentials compute $w_\ell$ shortest paths and are thus perfect.

@@

The most similar approaches to CH-Potentials are ALT and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
Just as our approach, ALT is A*-based.
However, ALT's potentials are not perfect. 
%
CPD-Heursitics are a combination of A* and Compressed Path Databases (CPD). 
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
In every potential evaluation, a path to the target is compute, whose length is used as potential.
The resulting potential is perfect.
Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a perfect shortest distance array with respect to $w_\ell$ as potential.
This is the tightest potential achievable during preprocessing. 
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

We introduce CH-Potentials in Section~\ref{sec:main-algo}. 
A two-phase technique to efficiently compute perfect potentials.
Combined with A*, this yields an optimal, efficient, and flexibile route planning algorithms.
Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
Beside the potentials, we describe, in Section\ref{sec:low-deg-improvment}, some improvments to A* to more efficiently process low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and showing how they can be solved using CH-Potentials.
Finally in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach. 
By comparing against Oracle-A*, we show that no alternative approach can be significantly faster.

@@

Using a CH, we can easily compute perfect A* potentials.
In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
The query phase consists of an A* search, where the evaluation of the potential $\pi(v)$ performs a CH-query from $v$ to $t$.
%
This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
Fortunatly, we can do better.
However, for this we need another component called PHAST \cite{dgnw-phast-13}.

@@

We introduce a novel way to compute potentials for A*.
The potentials are calculated utilizing a CH on a lower bound graph and perfect with regards to that lower bound graph.
With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.

>>>>

\pdfinfo{
/Title (A* with Fast and Tight Heuristic/Potentials in Road Networks)
/Author (Ben Strasser, Tim Zeitz)
/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
}
\title{A* with Fast and Tight Heuristic/Potentials in Road Networks}
\author{Ben Strasser\\
academia@ben-strasser.net\\
\And
Tim Zeitz\\
tim.zeitz@kit.edu\\
Karlsruhe Institute of Technology
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and optimal routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic, also called potential, function.
Computing a good heuristic function is a challenge of its own.
To avoid the need for a heuristic function, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
These techniques are not based on A* and therefore do not need a heuristic function.
They achieve speed and optimality but sacrifize A*'s flexibility.
In this paper, we use hierarchical techniques to compute heuristic functions.
We call our technique CH-Potential.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
\end{abstract}

@@

In this paper, we consider an algorithm, where the query phase consists of an A* search.
The preprocessing phase computes auxilary data to efficiently evaluate the heuristic/potential.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
Following, their terminology, we refer to the heuristic as \emph{potential} function $\pi$.
Further, paths are computed from a source node $s$ to a target node $t$.

We say that a potential $\pi$ is \emph{tight}, if no tighter potential can be computed in the preprocessing phase.
The shortest path distances with respect to $w_\ell$ are tight.
CH-Potentials compute $w_\ell$ shortest paths and are thus tight.

@@

The most similar approaches to CH-Potentials are ALT and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
Just as our approach, ALT is A*-based.
However, ALT's potentials are not tight. 
%
CPD-Heursitics are a combination of A* and Compressed Path Databases (CPD). 
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
In every potential evaluation, a path to the target is compute, whose length is used as potential.
The resulting potential is tight.
Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$ as potential, i.e., it is tight.
This is the tightest potential achievable during preprocessing. 
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

We introduce CH-Potentials in Section~\ref{sec:main-algo}. 
A two-phase technique to efficiently compute tight potentials.
Combined with A*, this yields an optimal, efficient, and flexibile route planning algorithms.
Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
Beside the potentials, we describe, in Section\ref{sec:low-deg-improvment}, some improvments to A* to more efficiently process low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and showing how they can be solved using CH-Potentials.
Finally in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach. 
By comparing against Oracle-A*, we show that no alternative approach can be significantly faster.

@@

Using a CH, we can easily compute tight A* potentials.
In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
The query phase consists of an A* search, where the evaluation of the potential $\pi(v)$ performs a CH-query from $v$ to $t$.
%
This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
Fortunatly, we can do better.
However, for this we need another component called PHAST \cite{dgnw-phast-13}.

@@

We introduce a novel way to compute potentials for A*.
The potentials are calculated utilizing a CH on a lower bound graph and tight with regards to that lower bound graph.
With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.
